HMM

### 2015-12-11 paper3.Rnw
miror edits

For movement behaviours that change on a very fast time scale, such
that movement behaviours at successive observations are independent,
\emph{finite mixture models} (FMM) -- a special case of HMMs where the
probability of state occupancy is independent of the previous state --
can adequately describe movement. \todo{are there animal movement
  studies that use FMMs?} 
%% \citep{tracey_mapping_2012}  
When movement varies over long time scales
(relative to the time between observations) with little short-term
persistence or correlation, movement could be well represented by FMMs
where the occupancy probabilities changed systematically
\todo{smoothly? deterministically?} over time.  Thus FMMs and HMMs,
with or without temporal variation in the occupancy or switching
probabilities, form a useful family of models for capturing
changes in movement behaviour over a range of time scales.

@@

\item Emission Probability: $P($\boldmath{$y_{k}$}\unboldmath $\mid z_{k},$\boldmath{$x_{k}$}\unboldmath$)$, a vector of observation density \boldmath{$y_{k}$} \unboldmath condition on covariates \boldmath{$x_{k}$} \unboldmath at state $z_{k}$ at time $t=k$.  
\end{enumerate}
The likelihood above is the likelihood of the observed sequence given a particular
hidden sequence. In order to calculate the likelihood of the observed sequence given the
HMM, we need to sum over all possible hidden sequences. This direct method is computati
nally inefficient, but there exist more efficient ways to computation the likelihood and
estimate parameters numerically \citep{zucchini_hidden_2009}.

@@

We simulated simple covariate dependent transitions and fitted with covariate independent transitions to confirm if the covariate independent transition HMM over fit the true number of states from the covariate dependent transition HMM. From simulation, it is likely to over estimate the number of states by fitting time homogenous transition HMMs if the true transition is time inhomogeneous. See Appendix.

@@

As a compliment, we also added FMM and FMM with temporal heterogeneities to compare the temporal effects in goodness of fit. Looking at figure 2 time inhomogenous transition HMMs (vertically) and 5 states yield the lowest $\bigtriangleup$ BIC. The time inhomogeneous transition HMMs reduced the number of states from the HMM by 1 state consistently across all panthers via $\bigtriangleup$ BIC. Two of the four panthers dropped to 4 states from 5 states after adding time inhomogeneity transition moving closer to a biologically and statistically interpretable level. Although adding the time inhomogeneous only reduce the number of states by 1, it does support that adding temporal heterogeneity reduce the number of states and at the same time, increase goodness of fit by BIC.

>>>>

For movement behaviours that change on a very fast time scale, such
that movement behaviours at successive observations are independent,
\emph{finite mixture models} (FMM) -- a special case of HMMs where the
probability of state occupancy is independent of the previous state --
can adequately describe movement \citep{tracey_mapping_2012}.  
When movement varies over long time scales
(relative to the time between observations) with little short-term
persistence or correlation, movement could be well represented by FMMs
where the occupancy probabilities changed deterministically over time.  Thus FMMs and HMMs,
with or without temporal variation in the occupancy or switching
probabilities, form a useful family of models for capturing
changes in movement behaviour over a range of time scales.

@@

\item Emission Probability: $P($\boldmath{$y_{k}$}\unboldmath $\mid z_{k},$\boldmath{$x_{k}$}\unboldmath$)$, a vector of observation density \boldmath{$y_{k}$} \unboldmath condition on covariates \boldmath{$x_{k}$} \unboldmath at state $z_{k}$ at time $t=k$.  
\end{enumerate}
The likelihood above is the likelihood of the observed sequence given a particular
hidden sequence. In order to calculate the likelihood of the observed sequence given the
HMM, we need to sum over all possible hidden sequences. This direct method is computationally inefficient, but there exist more efficient ways to computation the likelihood and
estimate parameters numerically \citep{zucchini_hidden_2009}.

@@

@@

As a compliment, we also added FMM, FMM with sinusoidal temporal heterogeneities, and HMM with hourly transitions to compare the temporal effects in goodness of fit. Looking at figure 2 time inhomogenous transition HMMs (vertically) and 5 states yield the lowest $\bigtriangleup$ BIC. The time inhomogeneous transition HMMs reduced the number of states from the HMM by 1 state consistently across all panthers via $\bigtriangleup$ BIC. Two of the four panthers dropped to 4 states from 5 states after adding time inhomogeneity transition moving closer to a biologically and statistically interpretable level. Although adding the time inhomogeneous only reduce the number of states by 1, it does support that adding temporal heterogeneity reduce the number of states and at the same time, increase goodness of fit by BIC.
